---
format:
  revealjs:
    theme: [night, assets/phdcn_style.scss]
    logo: img/phdcn_logo_white.svg
    incremental: false
    self-contained: false
    width: 1200
    height: 800
editor: source
mouse-wheel: true
history: false
---



## [Mid-life health and childhood exposure to dangerous contexts]{.r-fit-text} {background-image="img/stephan-cassara-KnAIsBuitGg-unsplash.jpg" background-opacity="0.3"}

&nbsp;

|                    |                           |
|-------------------:|:--------------------------|
| Charles C. Lanfear | *University of Cambridge* |
| David S. Kirk      | *University of Oxford*    |
| Robert J. Sampson  | *Harvard University*      |



```{r setup}
#| include: false
knitr::opts_chunk$set(dev = "ragg_png",  
                      dev.args = list(bg = 'transparent'),
                      message = FALSE,
                      echo = FALSE,
                      warning = FALSE,
                      fig.showtext = TRUE,
                      fig.retina = 5,
                      fig.align = "center")
library(tidyverse)
library(showtext)
plot_font <- "Open Sans"
font_add_google(name = plot_font)
showtext_auto()
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
```


## [Exposure]{background-image="img/stephan-cassara-KnAIsBuitGg-unsplash.jpg" background-opacity="0.1"}

Exposure to gun violence is pervasive

## [Danger and gun carrying]{background-image="img/stephan-cassara-KnAIsBuitGg-unsplash.jpg" background-opacity="0.1"}

Show SA paper

Exposure to dangerous context--including gun violence--predicts early-age carrying

... but most people age out... what are lasting consequences?

## [Handgun violence over the life course]{.r-fit-text}{background-image="img/stephan-cassara-KnAIsBuitGg-unsplash.jpg" background-opacity="0.1"}

Most concealed handgun research focuses on adolescents

. . .

But handgun offending occurs throughout the life course


. . .

We know little about **glater life consequences**


## Questions {background-image="img/clay-banks-nsAn3nSW5T0-unsplash.jpg" background-opacity="0.3"}

* **Physical health**
* **Mental health**



* Substance use
* Mortality
* Residential mobility

## {.nostretch background-image="img/clay-banks-nsAn3nSW5T0-unsplash.jpg" background-opacity="0.1"}

![](img/phdcn_logo_white.svg){width=700px align=center}

* Cohorts born 1981, 1984, 1987, and 1996
   * Average ages 25 to 40 in 2021

* Key measures:
   * **Mental Health**: Kessler K6 distress scale
   * Physical Health: 
      * **5 category self-reported**
      * Diagnosed serious health conditions
  
   * **Exposure**: Age first saw someone else shot

* Survey and attrition weights

# Physical Health {background-image="img/pedro-lastra-5Bv3EcijAl4-unsplash.jpg" background-opacity="0.3"}


# Continuity {background-image="img/benjamin-suter-mpLex62zVKQ-unsplash.jpg" background-opacity="0.3"}



# Mental distress {background-image="img/ozzie-stern-dkwJLowVvl4-unsplash.jpg" background-opacity="0.2"}

```{r, eval=FALSE}

legend_labs <- c("Early carry" = "Gun carry before age 21")
mh_pooled_data |>
  pivot_longer(c(early_carry, seen_shot)) |>
  mutate(name = str_to_sentence(str_replace(name, "_", " "))) |>
  ggplot(aes(x = mhealthsum_5, fill = value, group = value, color = value)) + 
  scale_fill_manual(values = c("yes" = "#F8766D", "no" = "#00BFC4")) +
  scale_color_manual(values = c("yes" = "#F8766D", "no" = "#00BFC4")) +
  facet_wrap(~name, labeller = labeller(fill =, color = )) +
  labs(color = NULL, fill = NULL) +
  geom_density(alpha = 0.5, bounds = c(0, 24)) +
  annotate("segment", x = 13, xend= 13, y = 0, yend = 0.15, color = "red") +
  annotate("segment", x = 5, xend= 5, y = 0, yend = 0.15, color = "yellow") +
  annotate("text", x = 13, y = 0.16, color = "red", label = "Severe") +
  annotate("text", x = 5, y = 0.16, color = "yellow", label = "Moderate") +
  geom_text(data = mh_pr_df, aes(x = x, y = y, label = pr, color = value), show.legend=FALSE) +
  theme_void() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.45, 0.85),
        strip.text = element_text(family = plot_font,  color = "white", size = rel(1.5)),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        text = element_text(family = plot_font,  color = "white"))
  
```

```{r}
load("data/import/plot_data.RData")
mh_preds |>
  mutate(color = as.numeric(estimate == max(estimate)), .by = variable) |>
  ggplot(aes(x = term, y = estimate, fill = color)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~variable, scales = "free_x") +
  labs(y = NULL) +
  theme_minimal() +
  theme(
    panel.spacing.y = unit(12, "pt"),
    axis.text = element_text(color = "white"),
        panel.grid = element_blank(),
        strip.text = element_text(family = plot_font,  color = "white", size = rel(1.5)),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        text = element_text(family = plot_font,  color = "white"))
```

## [Probability of carry onset by age period]{.r-fit-text}{background-image="img/ozzie-stern-dkwJLowVvl4-unsplash.jpg" background-opacity="0.1"}

```{r}
mh_preds_4 |>
  mutate(color = as.numeric(estimate == max(estimate)), .by = variable) |>
  ggplot(aes(x = term, y = estimate, fill = color)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~variable, scales = "free_x") +
  labs(y = NULL) +
  theme_minimal() +
  theme(axis.text = element_text(color = "white"),
        panel.grid = element_blank(),
        strip.text = element_text(family = plot_font,  color = "white", size = rel(1.5)),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        text = element_text(family = plot_font,  color = "white"))
```

# Physical health

```{r}
ph_preds |>
  # mutate(color = as.numeric(estimate == max(estimate)), .by = variable) |>
  ggplot(aes(x = outcome, y = estimate, group = term, fill = term)) +
  geom_col(show.legend = FALSE, position = "dodge") +
  facet_wrap(~variable, scales = "free_x") +
  labs(y = NULL) +
  theme_minimal() +
  theme(
    panel.spacing.y = unit(12, "pt"),
    axis.text = element_text(color = "white"),
        panel.grid = element_blank(),
        strip.text = element_text(family = plot_font,  color = "white", size = rel(1.5)),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        text = element_text(family = plot_font,  color = "white"))
```

## Takeaway{background-image="img/joel-mott-s-rsM-AktbA-unsplash.jpg" background-opacity="0.1"}


## New Questions {background-image="img/joel-mott-s-rsM-AktbA-unsplash.jpg" background-opacity="0.1"}


## Feedback and Questions {.smaller background-image="img/joel-mott-s-rsM-AktbA-unsplash.jpg" background-opacity="0.1"}

Contact:

| Charles C. Lanfear
| Institute of Criminology
| University of Cambridge
| [cl948\@cam.ac.uk](mailto:cl948@cam.ac.uk)

&zwj;

For more about the PHDCN+:

| PHDCN\@fas.harvard.edu
| <https://sites.harvard.edu/phdcn/>
| [https://doi.org/10.1007/s40865-022-00203-0](https://sites.harvard.edu/phdcn/)

# Appendix {background-image="img/max-bender-yb7Yg3Rv7WA-unsplash.jpg" background-opacity="0.3"} 
